GPU-based Adaptive Beamforming for Active Sonar Imaging

### 2012-05-07 ECUA2012/gpu_capon_ecua2012.tex
la2pdf commit - 2012:05:07	18:07:08	(Mon 7.May)

\usepackage{tikz}
\usetikzlibrary{shapes,snakes}
\usepackage{amsmath,amssymb}

@@

When designing an active sonar system one usually finds the image quality limited by the laterial resolution and contrast. This is a physical limitation related to the array geometry and the frequency of the transmitted signal.

The Minimum Variance Distortionless Response (MVDR, or Capon) beamformer is one such technique. While ensuring unit gain in the look direction, it computes the set of weights that minimizes the power accumulated by the array from other directions. This ability to suppress interference power improves the image detail level and \gls{SNR}, and the effect can be dramatic if the noise and interference is primarily focused around specific angles. The feasibility of using the \gls{MVDR} beamformer in active sonar imaging has been demonstrated in e.g.~\cite{Blomberg2012a,Blomberg2011,Dursun2009}.

@@

\newpage
\begin{figure}[!t]
\centering
\includegraphics[width=\linewidth]{gfx/img_holmengraa.pdf}
\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa.}\label{holmengraa}
\end{figure}

\section{Implementation}

@@

@@

\newpage

\section{Results}

@@

% A sidescan image of the 1500 dwt oil tanker wreck Holmengraa is shown in \Fig{holmengraa}.  depth~\cite{holmengraa}. The sidescan image was created using data from the HISAS 1030 sonar, which is rather unsuited for this purpose because of its large opening angle. This, and the fact that the wreck was imaged at a range of about 105\,m makes the image quality poor, but sufficient to compare our beamformers.

% % In the Holmengraa image we note again that the LCA produces a cleaner shadow and better edge definition. The MV method performed almost identically to LCA in this case, and was omitted.

@@

% down to these cores and back to the \gls{CPU} is also when copying data to and from the \gls{GPU} are also rather high, so each core should be   performing calculations there, and copying the data back to the \gls{CPU}
%  but in order to fully exploit this potential we must be aware of its limitations.  
% When adapting an algorithm to a massively parallel hardware such as \glspl{GPU}, one must be familiar with its limitations in order to that the potential that lies within these chips can not be fully realised 
% When trying to fit any algorithm to massively parallel hardware such as \glspl{GPU}, one must be careful to select a proper decomposition scheme for the device at hand. Modern \glspl{GPU} have several hundred computing cores, but only a limited amount of  are  various decomposition schemes should be contemplated before performing the implecareful decomposition of the 
% Let us first consider the \emph{computation of $\hat{\mat R}$}. Notice that when implementing the formula directly a lot of summations and multiplications will be repeated unnecessary. To illustrate this, consider for instance a case with $M=6$ sensors that are grouped into subarrays of length $L=3$ where all but one element overlap. Let us further assume that we need to average over 7 time samples, i.e. that $N_{\text{avg}} = 3$. 
% These procedures are commutative, so the order in which we perform them does not matter. But for the sake of this example let us start by performing the time averaging. Assuming that $\vec x_i$ and $\vec x_j$ have a zero mean, a time averaged covariance coefficient $r_{ij}$ in $\hat{\mat R}$ may be obtained by (omitting the normalisation):
% \begin{align*}
% r_{ij}[n] &= \sumb{n'=n-N_{\text{avg}}}{n+N_{\text{avg}}} x_i[n']\,x_j[n'],
% \end{align*}
% which can be easily rewritten in a sliding window fashion
% \begin{align*}
% \hat r_{ij}[n] &= r_{ij}[n-1] - x_i[n\text{-}N_\text{avg}\text{-}1]\,x_j[n\text{-}N_\text{avg}\text{-}1]
%                          + x_i[n\text{+}N_\text{avg}]\,x_j[n\text{+}N_\text{avg}].
% \end{align*}
% Hence, whenever $N_\text{avg}>1$ the sliding window approach will demand less instructions at the expense of a slightly higher memory consumption. The next thing to realise is that not all $r_{ij}[n]$ need to be computed. Since $r_{ji} = r^*_{ij}$, and the subarrays only span $L$ elements, we only compute an upper triangular band of R (marked \Green{green})
% \begin{figure}[H]
% % \begin{narrow}{-3.5cm}{-3.5cm}
% \includegraphics[width=\linewidth]{gfx/capon_build_R_full.eps}
% \caption{Building $\hat{\mat R}$: Symmetry and element coverage}\label{buildingRfull}
% % \end{narrow}
% \end{figure}
% The final step is to average all the subarray matrices to form $\hat{\mat R}$ by 
% {\renewcommand*{\arraystretch}{1.8}
% \begin{align*}
% \hat{\mat R}_\text{sub} &= {\color{Blue}\mat R_{\text{sub},1} + \mat R_{\text{sub},2} + \mat R_{\text{sub},3} + \mat R_{\text{sub},4}} \\
% &= \left[\;
% \begin{matrix}
% \color{Green}r_{00}+r_{11}+r_{22} & \color{Red}r_{01}+r_{12}+r_{23} & r_{02}+r_{13}+r_{24} \\
%                                        & \color{Green}r_{11}+r_{22}+r_{33} & \color{Red}r_{12}+r_{23}+r_{34} \\
%                                        &                                        & \color{Green}r_{22}+r_{33}+r_{44}
% \end{matrix}
% \;\right]
% \end{align*}}%
% Note that consecutive elements along the diagonals overlap on all but one covariance coefficient, hence a sliding window may also be used here. The subarray size $L$ is usually quite a bit larger than the time averaging window, so there is more to gain from implementing the sliding window here.\\\\
% {\bf Method complexity:}\\
% \begin{tabular}{l l}
% No optimizations            & $2M^2\,(2\Navg+1)$ \\
% Banded symmetric            & $L(2M-L+1)\,(2\Navg+1)$ \\
% Sliding, banded, symmetric  & $1.5L(2M-L+1)$ \\
% \end{tabular}\\\\

% \subsubsection{Computing $\text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a$}
% 
% There are two ways of finding $\text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a$, either by solving the equation directly or by inverting $\hat{\mat R}$ first. Solving the equation is less computationally intensive than inverting $\hat{\mat R}$, and seems like the obvious choice. However, with iterative methods for updating $\eRi$ for each range pixel the situation is changed...
% 
% 
% \section{Results/Discussion}
% 
% 
% 
% 
% 
% \todopar{Case:
% \begin{itemize}
% \item JOE
% \item Sonar. Channel count up to 32/64 ish channels?
% \item Other articles evaluates \gls{MVDR} image quality, and some have shown that speedups can be achieved by reducing the complexity of the traditional \gls{MVDR} algorithm or by implementing specific versions on a GPU.
% \item This article compare a few promising ways to implement \gls{MVDR} in general, in light of general complexity and suitability for implementation on a CPU and GPU.
% \end{itemize}
% }
% 
% \begin{itemize}
% \item As briefly as possible, introduce beamforming.
% \item Adaptive beamformer's potential lies in its ability to suppress interference power
% \item Why adaptive beamformers struggle in active sonar systems. Correlated noise, robustification kills the adaptive potential. Quite computationally intensive. Constraints must be applied in one way or another - parameters must be tuned.
% \item The three main ways to implement \gls{MVDR} \cite{\gls{MVDR}1969} is \todo{not entirely true?}
% \begin{itemize}
% \item Traditional way: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
% \item Iterative methods: Woodbury, Conjugated gradients
% \item Beamspace
% \item LCA (or other methods, as reference)
% \end{itemize}
% \item Choice of robustification parameters $L,M,\Delta$ largely impacts what the ``optimal'' way of implementing \gls{MVDR} 
% \item In most imaging applications 
% Outline:
% \begin{itemize}
% \item Speeding up: Build $\eR$ and solve $\w = \frac{\Ri\a}{\a\H\Ri\a}$. 
% \item Iterative methods: Woodbury (Sherman Morrison), Conjugated gradients - show performance of GPU performance.
% \item Woodbury: Must take the average over many samples in space and time - lots of (small) outerproducts. Can be used if not a lot of averaging is required.
% \item Beamspace
% \item LCA (or other methods, as reference)
% \end{itemize}
% \end{itemize}
%

% \ \\
% \IEEEPARstart{T}{o} form images from a modern phased array sonar system the received wavefield is usually recorded, and then postprocessed by a digital beamformer. The beamformer applies delays and weights to the sensor channels, the beamformer adjusts the arrays spatial response to focus at one pixel at a time.  such that signals emanating from regions of interest add constructively, while ensuring that noise and interference from other angles do not. 
% 
% The imaging capabilities of a modern phased array sonar system depend on physical attributes such element response and array geometry, the transmitted signal, as well as the beamforming method being used on transmission and reception. Beamforming is the concept of applying delays and weights to the sensors channels to steer the arrays response to points of interest.

% 
% 
% Outline:
% \begin{itemize}
% \item \gls{MVDR}'s resonse when applying robustification
% \item Choice of window functions makes little difference.
% \item Steering and mainlobewidths have outer bounds.
% \item Beamspace?
% \item Chosen window plots - what may they tell us? Variance intensity values when using various windows.
% \item Assymmetric windows?
% \end{itemize}
% 
% 
% \begin{align}
% z[n] = \sumb{m=0}{M-1} w_m[n]^*x_m[n-\Delta_m] = \w\H[n]\x[n-\Delta_m]
% \end{align}
% 
% 
% \section{Methods}
% 
% Basically, we are working with a practical implementation of the \gls{MVDR} beamformer that computes a set of weights $\vec w$ for every single sample $n$ by solving:
% \begin{gather*}
% \vec w[n] = \frac{\hat{\mat R}\,\!^{-1}[n]\vec a}{\vec a\H\hat{\mat R}\,\!^{-1}[n]\vec a} = \frac{\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}{\vec a\H\text{\raisebox{1.9pt}{$\vec\chi$}}[n]}
% \end{gather*}%
% where
% % \newcommand\X{\text{\raisebox{2pt}{$\vec\chi$}}}
% \begin{gather*}
% \text{\raisebox{1.9pt}{$\vec\chi$}} = \hat{\mat R}\,\!^{-1}\vec a \qquad\qquad\text{is the solution to}\qquad\qquad \hat{\mat R}\text{\raisebox{1.9pt}{$\vec\chi$}} = \vec a.
% \end{gather*}

% 
% , and maximum suppression of while ensuring that the beamformer digitally  before each of the pixels are estimated one at a time. The resolution and contrast of such a system will depend on the systems spatial response, which ideally should be narrow  be very sharp in the desired direction its ability to achieve  fundamental principle of forming a sonar image is to record the received wavefield, 
% 
% image quality of a phased array sonar imaging systems depend on  the choice of weights to apply to each of the sensors are crucial. 
% 
% A modern phased array imaging system may be thought of as a spatial filter. To achieve the best possible performance, the 
% 
% resolution and contrast 
% 
% Adaptive beamformers have only recently been introduced in active sonar imaging. For a while they were considered unsuited for this purpose because the backscattered signal is largely correlated with the 
% 
% 
%

%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results}
%\label{fig_sim}
%\end{figure}

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command, the
% \label for the overall figure must come after \caption.
% \hfil must be used as a separator to get equal spacing.
% The subfigure.sty package works much the same way, except \subfigure is
% used instead of \subfloat.
%
% \begin{figure*}[!t]
% \centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
% \label{fig_first_case}}
% \hfil
% \subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
% \label{fig_second_case}}}
% \caption{Simulation results}
% \label{fig_sim}
% \end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals use top floats exclusively.
% However, Computer Society journals sometimes do use bottom floats - bear
% this in mind when choosing appropriate optional arguments for the
% figure/table environments.
% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
% floats. This can be corrected via the \fnbelowfloat command of the
% stfloats package.
% 
% 
% 
% \section{Conclusion}
% The conclusion goes here.
% 
% 
% 
% 
% 
% % if have a single appendix:
% %\appendix[Proof of the Zonklar Equations]
% % or
% %\appendix  % for no appendix heading
% % do not use \section anymore after \appendix, only \section*
% % is possibly needed
% 
% % use appendices with more than one appendix
% % then use \section to start each appendix
% % you must declare a \section before using any
% % \subsection or using \label (\appendices by itself
% % starts a section numbered zero.)
% %

% 
\section{Acknowledgments}

>>>>

% \usepackage{tikz}
% \usetikzlibrary{shapes,snakes}
\usepackage{amsmath,amssymb}

@@

When designing an active sonar system one usually finds the image quality limited by the laterial resolution and contrast. This is a physical limitation related to the lateral array geometry and the frequency of the transmitted signal.

The Minimum Variance Distortionless Response (MVDR, or Capon) beamformer is one such technique. While ensuring unit gain in the look direction, it computes the set of weights that minimizes the power accumulated by the array from other directions. This ability to suppress interference power improves the image detail level and \gls{SNR}, and the effect can be dramatic if the noise and interference is primarily focused around specific angles. The feasibility of using the \gls{MVDR} beamformer in active sonar imaging has been demonstrated in e.g.~\cite{Blomberg2012a,Blomberg2011,Dursun2009,Lo2004}.

@@

\newpage
\section{Implementation}

@@

\begin{figure}[!t]
\centering
\includegraphics[width=\linewidth]{gfx/img_holmengraa.pdf}
\caption{HISAS sidescan sonar (SSS) image of the shipwreck Holmengraa.}\label{holmengraa}
\end{figure}

@@

\newpage
\section{Results}

@@

@@

{\Large\bf\MakeUppercase Acknowledgments}
